visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}

### 2025-05-10 jss-article-rnw/article.Rnw
Started restructuring article.

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@

@@

\title{visualFields 1.x: Tools for Analyzing the Field of Vision in \proglang{R}}
\Plaintitle{visualFields 1.x: Tools for Analyzing the Field of Vision in R}
\Shorttitle{visualFields 1.x in \proglang{R}}

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

To install the released version of the package, run:
\begin{CodeChunk}
\begin{CodeInput}
install.packages("visualFields")
\end{CodeInput}
\end{CodeChunk}

To install the newest version of the package, run:
\begin{CodeChunk}
\begin{CodeInput}
install.packages("devtools")
library(devtools)
install_github("imarinfr/vf1/source")
\end{CodeInput}
\end{CodeChunk}

To load the package, use:
<<>>=
library(visualFields)
@

@@

@@

\subsection{Getting started with the package}
\subsection{Loading data}
The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).

@@

\subsection{Plotting the visual field}
The most straightforward plots show the sensitivities at the corresponding locations. These representations are very exact, but may be difficult to interpret. For learning purposes, here, we manually plot some visual field data.

vfisvalid(example2)

example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
                                 y = getlocmap()$coord$y, 
                                 td = as.vector(unlist(example2[,getvfcols()])))

ggplot(example2_dataframe, aes(x = x, y = y)) +
  geom_text(aes(label = td)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_fixed() +
  theme_void()

Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis. In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.

OD_dataframe <- data.frame(x = getlocmap()$coord$x, 
                             y = getlocmap()$coord$y, 
                             loc_label = paste0("l", 1:54),
                             is_blindspot = FALSE,
                             eye = "OD")

# Note the minus in the x variable for the left eye
OS_dataframe <- data.frame(x = -getlocmap()$coord$x, 
                             y = getlocmap()$coord$y, 
                             loc_label = paste0("l", 1:54),
                             is_blindspot = FALSE,
                             eye = "OS")

OD_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
OS_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE

ODOS_dataframe <- rbind(OS_dataframe, OD_dataframe)

# Use factor in order to show left eye on the left side
ODOS_dataframe$eye <- factor(ODOS_dataframe$eye, 
                             levels = c("OS", "OD"), 
                             ordered = TRUE)

ggplot(ODOS_dataframe, aes(x = x, y = y)) +
  coord_fixed() +
  xlim(-30, 30) +
  ylim(-24, 24) +
  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
  scale_color_manual(values = c("black", "red")) +
  facet_wrap(~ eye)

The visualField package provides functions for more sophisticated plotting, which will be explained later. \textbf{These functions automatically take care that right eyes and left eyes are plotted correctly.}

@@

\subsection{Identifying defective locations on the visual field}

@@

\subsection{Normative dataset and normative references}
The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.

@@

\subsubsection{Age linear models and mean normal visual fields}
The normative values and references stored in `nv` contains the coefficients of the age linear models:

@@

\subsection{Total deviation}
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).

@@

\subsection{Pattern deviation}
Pattern deviation can be calculated after obtaining the general height.
<<gh>>=
gh <- getgh(td)
print(gh)
pattern_defects <- td[, getvfcols()[1:10]] - gh
print(pattern_defects)
@

@@

\subsection{Total and Pattern deviation probabilities}
As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time. Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.

@@

[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters

It is possible to print sensitivities (with darker representing locations with lower sensitivities)

vfplot(example2, type = "s")

Pattern deviation values

vfplot(example2, type = "pd")

And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.

vfplot(example2, type = "tds")

And a similar hybrid plot, but with pattern deviation values and probabilities.

vfplot(example2, type = "pds")

@@

locmap <- locmaps$p10d2
gcmap <- vf2gc(locmap$coord)
par(mfrow = c(1, 2))
plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "10-2 grid of test locations")
lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
for(r in c(2.5, 5, 7.5, 10))
  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "Corresponding ganglion cells")
lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
for(r in c(2.5, 5, 7.5, 10))
  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
@

@@

The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012]. First, we can obtain the angle of incidence on the optic never head (ONH)
<<>>=
print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
@

@@

\subsection{Plotting progression}
The sparkline plot only shows the raw data: the series of visual field data (by default total deviation values) over time at each location of the visual field. The lines are plotted in red if they are very "noisy", i.e., if the median absolut deviation of the residuals from simple linear regression are greater than a certain threshold `thr` (by default 2 dB). Here is an example with the threshold set at 4 dB.

vfsparklines(example3, thr = 4)

The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion. By default it compares the first 3 visual fields against the last 3 of a series. And by default it compares total deviation values (and probabilities of damage, which relies on population based statistics). This example, however, shows sensitivities.

vflegoplot(example3, type = "s")

The outer polygon shows the grayscale for the average sensitivity over the first 3 visual fields, whereas the inner circles show the grayscale for the average of the last 3. The numbers inside the inner circles are the difference in sensitivity between the last 3 and first 3 visual fields.

@@

@@

\subsection{Acknowledgments}
The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large. Huge thanks to:
  - William H Swanson
  - Michael Wall
  - Andrew Turpin
  - Paul H Artes
  - Giovanni Montesano
  - Mitchell W Dul

\begin{leftbar}
This section can include details about the methods implemented in the \pkg{visualFields} package, such as statistical models, algorithms, or visualization techniques. For now, this section is left as a placeholder.
\end{leftbar}

\begin{leftbar}
This section can include results obtained using the \pkg{visualFields} package, such as example analyses or visualizations. For now, this section is left as a placeholder.
\end{leftbar}

@@

>>>>

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@

@@

\title{visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}}
\Plaintitle{visualFields 1.0.7: Tools for Analyzing the Field of Vision in R}
\Shorttitle{visualFields 1.0.7 in \proglang{R}}

@@

\begin{document}
%%\SweaveOpts{concordance=TRUE}

@@

@@

library(visualFields)
library(plotrix)

@@

The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:

<<eval=FALSE>>=

install.packages("visualFields")

To install the newest version of the package, run:

<<eval=FALSE>>=

library(devtools)
install_github("imarinfr/vf1/source")

\subsection{Loading data}
The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).

@@

@@

\subsubsection{Identifying defective locations on the visual field}

@@

\subsubsection{Normative dataset and normative references}
The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.

@@

\subsubsubsection{Age linear models and mean normal visual fields}
The normative values and references stored in `nv` contains the coefficients of the age linear models:

@@

\subsubsection{Total deviation}
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).

@@

\subsubsection{Pattern deviation}
Pattern deviation can be calculated after obtaining the general height.
<<gh>>=
gh <- getgh(td)
print(gh)
pattern_defects <- td[, getvfcols()[1:10]] - gh
print(pattern_defects)
@

@@

\subsubsection{Total and Pattern deviation probabilities}
As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time. Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.

@@

@@

The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters.

@@

A gallery with all available visual field plots is available online at \url{
https://rpubs.com/huchzi/1307927}.

@@

\subsection{Plotting progression}

The plot used for showing progression are shown in the gallery, which can be found at: \url{https://rpubs.com/huchzi/1307927}.

@@

locmap <- locmaps$p10d2
gcmap <- vf2gc(locmap$coord)

<<echo=TRUE, fig=TRUE>>=

plot_grid <- function() {
  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
  for(r in c(2.5, 5, 7.5, 10))
    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
}

par(mfrow = c(1, 2))
plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "10-2 grid of test locations")
plot_grid()
points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")

plot(x = locmap$coord$x, 
     y = locmap$coord$y, 
     type = "n", 
     asp = 1,
     xlim = c(-10, 12),
     ylim = c(-10, 10),
     yaxt = 'n',
     xaxt = 'n',
     bty = 'n',
     xlab = "",
     ylab = "",
     main = "Corresponding ganglion cells")
plot_grid()
points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
@

The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012]. First, we can obtain the angle of incidence on the optic never head (ONH)

print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))

@@

@@

%% -- Acknowledgements ---------------------------------------------------

\section{Acknowledgments}
The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large. Huge thanks to:
  - William H Swanson
  - Michael Wall
  - Andrew Turpin
  - Paul H Artes
  - Giovanni Montesano
  - Mitchell W Dul
